---
title: "Optimbucket: Usage Example"
author: "Felipe Gerard"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How-to guide for package optimbucket.}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


The `optimbucket` package was written to manipulate ROC curves even in the presence of special values. Its original objective was to discretize continuous variables in a way that maximizes their Gini Index (or equivalently their AUC). There are also functions to manipulate the buckets manually. It can also help you measure variable performance in different metrics such as AUC, Kolmogorov-Smirnov statistic and Information Value. I wrote this vignette to illustrate what the package can do.

Let's start with loading the required libraries and generating some random data:

```{r, message=FALSE}
require(dplyr)
require(tidyr)
require(ggplot2)
require(optimbucket)

ng <- 10000
nb <- 1000

goods <- data.frame(
  x = rnorm(ng, mean = ifelse(runif(ng) > 0.2, -2, 1)),
  y = 0
)
bads <- data.frame(
  x = rnorm(nb),
  y = 1
)

d <- rbind(goods, bads) %>%
  mutate(y = factor(y),
         z = 2*x + 1 + rnorm(ng+nb, 0, 5),
         w = round(x - z  - 1 + rnorm(ng+nb, 0, 1)),
         w = ifelse(runif(ng+nb) < 0.01, # Adding some special values to w
                    -999,
                    ifelse(runif(ng+nb) < 0.01,
                           -998,
                           w)))
head(d)
```

We can look at the separation each variable variable induces between the classes:

```{r, echo=F, fig.width=8, fig.height=4}
d %>%
  filter(w > -998) %>%
  gather(key, value, x, w, z) %>%
  ggplot(aes(value, fill=y)) +
  geom_density(alpha=0.5) +
  facet_wrap(~ key)
```

# Building ROC Curves

The first step is to create a ROC curve for each variable. The `ngroups` argument controls the level of detail of the ROC curve (`NULL` for all the detail); the more detail, the more expensive the ROC curve. I recommend no more than 300 groups. This way you can also control the size of the smallest bucket: 100/ngroups % of the non-special population will be the smallest bucket possible when we optimize later on. `level.bad` simply corresponds to the level corresponding to "event". The `special.values` argument can either be a vector containing special values common to all the variables or a named list with such a vector for each variable.

```{r}
wrs <- wroc(y ~ x + z + w, d, ngroups = 50, level.bad = 1,
            special.values = list(w = c(-998, -999)))

# Note: if you want to create a ROC curve for all the variables, you can use the following formula: y ~ .
```

This yields a `wroc.list` object, which is simply a named list of `wroc` objects we can later manipulate.

**Hint:** You can always build a single ROC curve:

```{r}
wr <- wroc(d$x, d$y, ngroups=20)
```

**Hint:** You can select a few of the ROC objects using integer indexes or the variable names with standard notation: `wrs[['w']]`. You can also join two or more lists with the combination function `c` like: `c(wrs['w'], wrs[c('x','z'])` (note the single brackets here).

# Visualizing ROC Curves

You can visualize detailed aspects of a variable using the standard `print` function and also some more facts using `summary`.

```{r}
print(wrs)
ss <- summary(wrs)
print(ss)
```

You can also use the dedicated `plot` function to visualize the ROC curves in several ways. The types 'roc' and 'accum' (the default) show a ROC curve and an equivalent of the ROC curve using false negative rate versus true negative rate, respectively. The types 'trend' and 'woe' show the population in each bucket along with either the probability of being in the positive class ('trend') or the Weight of Evidence ($-log(p/(1-p))$).

```{r, fig.width=7, fig.height=4}
plot(wr)
plot(wrs$w, 'woe')
plot(wrs$z, 'roc')
```

# Optimizing ROC Curves

To optimize a ROC curve or each ROC curve of a list individually, you can call `optimze`. The result is again a ROC curve with some additional info about which buckets were pasted. Note that for single curves you can choose the trend used, but for lists only the 'auto' option is available. You can pick individually by manipulating the list manually.

```{r}
wr2 <- optimize(wr, trend='auto')
wr3 <- optimize(wr, trend='lower')
wr4 <- optimize(wr, trend='upper')
owrs <- optimize(wrs)
oss <- summary(owrs)

print(owrs)
print(oss)
```

You can also optimize the ROC curves subject to a specific number of final buckets with `analyze`:

```{r, fig.width=7, fig.height=4}
owr <- analyze(wr, nbuckets=3, trend='auto')

summary(owr)
plot(owr, 'woe')
```

# Manual Bucketing

You can bucket manually either instead or as well as optimizing automatically. You pass a list of buckets and they get pasted to the right:

```{r, fig.width=7, fig.height=4}
plot(wr, 'woe')
wa <- subset(wr, c(2:3, 14:19))
plot(wa, 'woe')
```


# Transforming Raw Variables into WoE Form

You can transform the original variables into WoEs, bucket numbers, probability of a bucket, etc, using `predict` in the standard way. For single curves the return value is a vector; for lists it is a `data.frame` including the transformed variables and optionally the original data.

```{r}
x_woe_vec <- predict(wr2, d$x, 'woe')
head(x_woe_vec)
all_woes <- predict(owrs, d, 'woe', keep.data=F)
head(all_woes)
all_woes_with_data <- predict(owrs, d, 'woe', keep.data=T)
head(all_woes_with_data)
```

# Exporting ROC Curves

You can export the information tables contained within `wroc` objects using `write.table`, but I wrote some functions to make it easier. If you don't specify the file name, the information tables get copied to the clipboard for easy pasting into Excel, for example.

```{r, eval=F}
# Save ROC curves to file
copy(owrs, export = T, file = 'example_ROC_curves.tab')
# Copy ROC curves to clipboard
copy(owrs)
# Copy basic stats to clipboard
copy(ss)
```



# Miscellaneous

* You can reset bucket numbers after optimizing using `reset.buckets`.
* You can just get the performance of (lists of) curves using `performance`.
* You can also use `performance` to calculate AUC and Gini indexes of variables without constructing a `wroc` object first.







